Predicting prolonged apneic episodes using machine learning during nurse-administered procedural sedation

### 2021-03-16 manuscript.Rmd
multi plots

For the aggressive strategy, only triggering an alarm for apneas predicted to be prolonged would reduce the total alarm burden and potentially reduce the risk of alarm fatigue. The downside of applying a machine learning approaching to the aggressive apnea treatment strategy would be that some prolonged apneas may not receive early intervention if the model incorrectly predicted that the apnea will not last for >30 seconds (i.e. the false negatives). For the conservative strategy, triggering an alarm at the 15-second timepoint for apneas predicted to be prolonged could reduce the total time of apnea, as treatment could be initiated earlier. The downside of applying a machine learning approaching to the conservative apnea treatment strategy would be that there is the potential that it could increase the total alarm burden if the ratio of false positives (predicting the apnea will last for >30 seconds but it does not) to true positives (correctly predting at the 15 second timepoint that the apnea will last for >30 seconds) is high. The aim of this study was to determine the accuracy of machine learning models for predicting at the 15 second timepoint if a period of apnea will persist for 30 seconds or more with the goal of determining whether operationalizing these predictions in practice as alarm triggers would be beneficial.

@@

The prediction goal was to classify apneic events at the 15-second time-point as being either short (i.e. terminated prior to 30 seconds) or prolonged (persisted for 30 seconds or longer). The prediction algorithm was compared against typical default alarm settings for capnography monitors. These are the aggressive strategy, which triggers an alarm after 15 seconds of apnea, and the conservative strategy, which triggers an alarm after 30 seconds of apnea.

@@

Several raw demographic (age, sex) and clinical (ASA score, diagnosis of sleep apnea, dose and type of sedation and analgesia administered) variables were used as predictors. Features related to sedation dosing used as predictors in the model were the total dose of sedation and number of doses of sedation administered, time since first sedation and time since the previous dose of sedation. Other features were extracted from the capnography waveform to use as predictors, such as the previous respiratory state (defined as either normal breathing or abnormal), duration of the previous apneic event, time since the previous apneic event and total number of apneic events. A total of 18 variables were used.

@@

Analyses were performed using R[@rcore2021]. Data as well as details about how to access the code and a reproducible computing environment required to verify the results is available from this webpage: <https://github.com/awconway/capnopred>. Split into a training and test set containing 75% and 25% of the data, respectively.

The best model was chosen with respect to the optimal balance between maximising the ability to predict prolonged events and alarm frequency. The alarm frequency metric used was the ratio of correct predictions of short apneic periods to the total number of apneic periods that triggered alerts, which was calculated as TP/(TP + FP + apneas > 30seconds).

#### Modelling
Pre-processing steps included normalizing numeric predictors and using an interaction term for the duration of the previous respiratory state and total number of apneic events.
- Normalized predictors
- interaction terms

@@

We used the net benefit decision analytic measure to assist with deciding whether using the models in practice would lead to better outcomes on average than the default capnography alarm management strategies currently used in practice. The default strategies are: 1) the aggressive approach, which involves triggering an alarm after brief periods of apnea (typically 15 seconds); and 2) the concervative approach, which involves triggering an alarm for only prolonged periods of apnea (typically 30 seconds). The net benefit is calculated using the formula:

@@

Where $NB$ is the net benefit, $TP$ is the total number of true positives (apneic event at 15 seconds which was predicted to be prolonged and did persist for >30 seconds), $FP$ is the total of number of false positives (apneic events at 15 seconds that was predicted to be prolonged but did not persist for >30 seconds), $N$ is the sample size and $Pt$ is the probability threshold.

This formula essentially transforms the total number of true positives and false positives into a standardized scale, weighted by the relative harm or a false-positive result.[@vickers2006decision] For example, a net benefit of 0.07 means the net benefit of using the model would be 7 true positives from every 100 predictions from the model. This net benefit can result from any combination of true positives and false positives.[@vickers2016net] The higher the value of $Pt$, the more importance is placed on the number of false positives.

A decision curve was created to plot net benefits across a range of probability thresholds. The decision curve was compared against the aggressive alarm management strategy (alarm triggered at 15 seconds of apnea) and the conservative alarm management strategy (alarm triggered at 30 seconds of apnea). The decision curve takes into account a plausbile range of clinician preferences for the point at which an intervention should be initiated to treat apnea (this could be, for example, something as simple as verbal stimulation). We tested thresholds in the range from 0.4 to 0.8. In a practical sense, this means that we decided that all clinicians would not ever accept a probability for prolonged apnea lower than 0.4 as a useful trigger for intervention, and, conversely, that all clinicians would always consider that intervention is indicated if the probability of prolonged apnea was higher than 0.8. These probability thresholds can be interpreted as value preferences in the context of decision curve analysis. For example, a clinician who is 'risk-averse' may elect for a more conservative probability threshold (closer to 0.4). Individual participant characteristics will also influence clinicians' decisions about probability thresholds. A clinician may elect to intervene when the probability for prolonged apnea is 0.5 for an elderley patient with multiple comorbidities but not for a young patient who may be more likely to be able to tolerate longer periods of apnea.

A threshold of 0.5 inherently indicates that avoiding a false positive is as valuable of an outcome as identifiying a true positive. Preferences for probability thresholds below 0.5 are weighted such that identifying a true positive is more valuable than avoiding a false positive. Preferences for probability thresholds above 0.5 are weighted such that avoiding a false positive is more valuable than identifying a true positive.  Decision curves can be interpreted such that the strategy with the highest net benefit at each probability threshold has the highest clinical value.[@vickers2016net]

@@

### Model developmenent and evaluation

Out of sample prediction accuracy for the models using cross-validation is summarized in Table 1. The best model, in terms of

@@

Net benefit associated with the random forest model exceeded the strategy of intervening for all patients (the default strategy for the 'aggressive alarm management approach') across all probability thresholds from the range of 0.4 to 0.8 (Figure 2A). The interpretation is that the best clinical outcome would be achieved for clinicians who are willing to initiate an intervention to treat apnea at the 15 second mark if the probability of it being prolonged was more than 40%, by using the random forest model. The net benefit associated with the random forest model was higher than the strategy of not intervening for any patients at the 15 second mark (the default strategy for the 'conservative alarm management approach') across all probability thresholds from the range of 0.4 to 0.8 (Figure 2B).

@@

Results from prior research indicated that using information about the history of previous respiratory states may be a promising solution in regards to predicting durations of apneic periods. It was found in a study of capnography waveform abnormalities during nurse-administered sedation that the hazard for apnea increased two-fold (HR 2.14, 95% CI 1.75 - 2.62) when a patient was in a state of hypoventilation (defined as >10% reduction in ETCO~2~ from baseline).[@conway2019pre] The risk of apnea also increased with each additional sedation dose (HR 2.86; 95% CI 2.15 - 3.81).[@conway2019pre] Results from an earlier study in a different population also supported the observations that the onset of apneic periods during sedation are associated with the history of previous respiratory states. Krauss and colleagues[@kraussCharacteristicsPredictorsApnea2016] used survival analysis to model the time to first apneic events in a sample of 312 patients undergoing procedural sedation with propofol or ketamine in the emergency department. It was identified that having an abnormal ETCO~2~ measurement 30, 60 and 90 seconds prior to an apneic event increased risk for apnea (hazard ratios with 95% CI for the different lag durations of 2.45 (1.63 - 3.69), 1.88 (1.21 - 2.92) and 2.06 (1.36 - 3.11), respectively).

We leveraged these preliminary findings about the associations between apneic events and the history of previous respiratory states in the present study by building a predictive model using a machine learning approach.

Using the machine learning model would produce a better clinical outcome, in terms of the balance between true positives and false positives for detection of apneic periods over 30 seconds, than a default strategy of intervening for everyone who has an apneic period ≥15 seconds.

>>>>

For the aggressive strategy, only triggering an alarm for apneas predicted to be prolonged would reduce the total alarm burden and potentially reduce the risk of alarm fatigue. The downside of applying a machine learning approaching to the aggressive apnea treatment strategy would be that some prolonged apneas may not receive early intervention if the model incorrectly predicted that the apnea will not last for >30 seconds (i.e. the false negatives). For the conservative strategy, triggering an alarm at the 15-second timepoint for apneas predicted to be prolonged could reduce the total time of apnea, as treatment could be initiated earlier. The downside of applying a machine learning approaching to the conservative apnea treatment strategy would be that there is the potential that it could increase the total alarm burden if the ratio of false positives (predicting the apnea will last for >30 seconds, but it does not) to true positives (correctly predicting at the 15-second timepoint that the apnea will last for >30 seconds) is high. This study aimed to determine the accuracy of machine learning models for predicting at the 15-second timepoint if a period of apnea will persist for 30 seconds or more with the goal of determining whether operationalizing these predictions in practice as alarm triggers would be beneficial.

@@

The prediction goal was to classify apneic events at the 15-second timepoint as being either short (i.e. terminated prior to 30 seconds) or prolonged (persisted for 30 seconds or longer). The prediction algorithm was compared against typical default alarm settings for capnography monitors. These are the aggressive strategy, which triggers an alarm after 15 seconds of apnea, and the conservative strategy, which triggers an alarm after 30 seconds of apnea.

@@

Several raw demographic (age, sex) and clinical (ASA score, diagnosis of sleep apnea, body mass index, dose and type of sedation and analgesia administered) variables were used as predictors. Features related to sedation dosing used as predictors in the model were the total dose of sedation and number of doses of sedation administered, time since first sedation and time since the previous dose of sedation. Other features were extracted from the capnography waveform to use as predictors, such as the previous respiratory state (defined as either normal breathing or abnormal), duration of the previous apneic event, time since the previous apneic event and total number of apneic events. A total of 18 predicotr variables were used.

@@

Analyses were performed using R[@rcore2021]. Data as well as details about how to access the code and a reproducible computing environment required to verify the results is available from this webpage: <https://github.com/awconway/capnopred>.

#### Modelling
We selected several candidate models to evaluate, including a random forest model, generalized linear model (logistic regression), lasso regression, ridge regression and the XGBoost model. Out of sample accuracy of the models was calculated using 10-fold cross-validation.

The best model was chosen with respect to the optimal balance between maximising the ability to predict prolonged events and alarm frequency. The alarm frequency metric used was the ratio of correct predictions of short apneic periods to the total number of apneic periods that triggered alerts, which was calculated as TP/(TP + FP + apneas > 30seconds). Pre-processing steps included normalizing numeric predictors and using an interaction term for the duration of the previous respiratory state and total number of apneic events.

@@

We used the net benefit decision analytic measure to assist with deciding whether using the models in practice would lead to better outcomes on average than the default capnography alarm management strategies currently in place. The default strategies are: 1) the aggressive approach, which involves triggering an alarm after brief periods of apnea (typically 15 seconds); and 2) the concervative approach, which involves triggering an alarm for only prolonged periods of apnea (typically 30 seconds). The net benefit is calculated using the formula:

@@

Where $NB$ is the net benefit, $TP$ is the total number of true positives (apneic event at 15 seconds which was predicted to be prolonged and did persist for >30 seconds), $FP$ is the total of number of false positives (apneic events at 15 seconds that was predicted to be prolonged but did not persist for >30 seconds), $N$ is the sample size (number of predictions made) and $Pt$ is the probability threshold.

This formula essentially transforms the total number of true positives and false positives into a standardized scale, weighted by the relative harm of a false-positive result.[@vickers2006decision] For example, a net benefit of 0.07 means the net benefit of using the model would be 7 true positives from every 100 predictions from the model. This net benefit can result from any combination of true positives and false positives.[@vickers2016net] A probability threshold of 0.5 indicates that avoiding a false positive is as important to a clinician as identifiying a true positive. Preferences for probability thresholds below 0.5 are weighted such that identifying a true positive is more valuable than avoiding a false positive. Preferences for probability thresholds above 0.5 are weighted such that avoiding a false positive is more valuable than identifying a true positive. For example, for a $Pt$ of 0.75, the 'value' of a false positive is three true positives (0.7/0.25 = 3). In other words, to create a net benefit from using the model at this probability threshold, there must be more than three true positives for every false positive prediction made from the model. Conversely, for a $Pt$ of 0.25, the 'value' of a false positive is weighted far lower, at only a third of a true positive (0.25/0.75 = 0.33). This means that a net benefit would be achieved if there was more than one true positive for every three false positives.  Decision curves can be interpreted such that the strategy with the highest net benefit at each probability threshold has the highest clinical value.[@vickers2016net]

We created a decision curve to plot net benefits across a range of probability thresholds for the aggressive alarm management strategy (alarm triggered at 15 seconds of apnea) as well as for the conservative alarm management strategy (alarm triggered at 30 seconds of apnea). The decision curve takes into account a plausbile range of clinician preferences for the point at which an alarm should be triggered to signal that the patient is apneic. We tested thresholds in the range from 0.4 to 0.5 for the agressive alarm management strategy. In a practical sense, this means that we decided that all clinicians who usually use the aggressive alarm strategy would not ever accept a probability for prolonged apnea lower than 0.4 as a useful alarm trigger, because there would be little difference between this strategy and just setting the alarm for all apneas. We also decided that all clinicians would always consider that an alarm is indicated for the aggresive alarm manamagement strategy if the probability of prolonged apnea was higher than 0.5. A range of values were used because these probability thresholds can be interpreted as value preferences in the context of decision curve analysis. For example, a clinician who is more 'risk-averse' may elect for a more conservative probability threshold (closer to 0.4). Individual participant characteristics will also influence clinicians' decisions about probability thresholds. A clinician may elect to intervene when the probability for prolonged apnea is 0.4 for an elderley patient with multiple comorbidities but not for a young patient who may be more likely to be able to tolerate longer periods of apnea. For the conservative alarm management strategy comparison, we chose to plot the range of probability thresholds from 0.7 to 0.8. Higher values were chosen because the number of false positives would be an important consideration for clinicians already using a conservative alarm management approach.

@@

### Discrimination

A plot of the area under the receiver operating characteristics curves for the models using predictions from 10-fold cross-validation is presented in Figure 1. The random forest  had the best discriminatory power of the models, with a mean AUROC score of 0.66 (SE 0.03). A threshold performance plot, which summarises the discriminatory power for the models, including values for sensitivity, specificity, positive predictive value and negative predictive value across all probability thresholds, is presented in Figure 2.

### Calibration

The random forest model had the best callibration. It approximated observed risk at moderate (0.5) to high (0.8) thresholds (Figure 3), although risk was overestimated at very low thresholds and slighlty underestimated between 0.4 to 0.5. Other models severely overestimated risk at low probability thresholds and underestimated risk at high probability thresholds.

@@

As the random forest performed the best in terms of discrimination and callibration, we chose this model to evaluate using decision curve analysis. Net benefit associated with the random forest model exceeded the aggressive alarm management strategy across all probability thresholds from the range of 0.4 to 0.5 (Figure 4A). The interpretation is that the best clinical outcome would be achieved for clinicians who are willing to initiate an intervention to treat apnea at the 15 second mark if the probability of it being prolonged was more than 40% by using the random forest model. The net benefit associated with the random forest model was higher than the conservative alarm management strategy across all probability thresholds from the range of 0.7 to 0.8 (Figure 4B).

@@

Results from prior research indicated that using information about the history of previous respiratory states may be a promising solution in regards to predicting durations of apneic periods. It was found in a study of capnography waveform abnormalities during nurse-administered sedation that the hazard for apnea increased two-fold (HR 2.14, 95% CI 1.75 - 2.62) when a patient was in a state of hypoventilation (defined as >10% reduction in ETCO~2~ from baseline).[@conway2019pre] The risk of apnea also increased with each additional sedation dose (HR 2.86; 95% CI 2.15 - 3.81).[@conway2019pre] Results from an earlier study in a different population also supported the observations that the onset of apneic periods during sedation are associated with the history of previous respiratory states. Krauss and colleagues[@krauss2016] used survival analysis to model the time to first apneic events in a sample of 312 patients undergoing procedural sedation with propofol or ketamine in the emergency department. It was identified that having an abnormal ETCO~2~ measurement 30, 60 and 90 seconds prior to an apneic event increased risk for apnea (hazard ratios with 95% CI for the different lag durations of 2.45 (1.63 - 3.69), 1.88 (1.21 - 2.92) and 2.06 (1.36 - 3.11), respectively).

We leveraged these preliminary findings about the associations between apneic events and the history of previous respiratory states in the present study by building a predictive model using a machine learning approach. A random forest model had the best discriminitive ability and calibration from those that we tested. However, it should be noted that the accuracy of this random forest model was still quite low (AUROC = 0.66). Additional research should be undertaken with larger sample sizes to validate our initially promising findings.

It should be noted, though, that many prediction modelling studies focused on clinical outcomes have yielded similarly low AUROC scores. For example, a recent study of the predictive ability of vital sign parameters for detection of clinical deterioration in subacute care patients yielded a AUROC score of 0.57.[@considine2020vital] When presented with a model with a low AUC scores, decision curve analysis can help elicit whether the model is "good enough" to use in practice. Based on our results, clinicians using the conservative alarm management strategy (alarm only at 30 seconds of apnea) who are willing to 'value' a false positive about 2-3 times more than a true positive would derive an overall net benefit from using the random forest model as a trigger for apnea alarms. This is because using the random forest model would produce a better outcome, in terms of the balance between true positives and false positives, for detection of apneic periods over 30 seconds, than a default strategy of waiting to trigger the alarm until the 30-second threshold.

### Limitations

Cross-validation

Small sample size

Convenience sampling from two sites.
